An Empirical Analysis of the Role of Energy in Economic Growth

### 2014-12-25 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Final edits to include Martin and my interpretation.

We find it beneficial to discuss results through a series of questions 
about the techniques presented above.

@@

>>>>

We find it beneficial to discuss results through a series of questions, 
answers to which are illuminated by the techniques and results presented above.

@@

\begin{table} \caption{Elasticities of substitution (UK).} \label{tab:sigma_UK} 
  \begin{center}
    \begin{tabular}{r l c l c l c} 
      \toprule
         & \multicolumn{2}{c}{$k$ \& $l$} & \multicolumn{2}{c}{$k$ \& $e$} & \multicolumn{2}{c}{$l$ \& $e$} \\
      \midrule
      \multirow{2}{*}{$\sigma_1$} & ($\underline{kl}$) & C 
                                        &  & 
                                              & &  \\
                                  & ($\underline{kl}$)$e$ & $\sim\sim\sim$ 
                                        & ($\underline{ek}$)$l$ & C
                                              & ($\underline{le}$)$k$ & C \\
      \midrule
      \multirow{3}{*}{$\sigma$} & &
                                      & ($\underline{k}l$)$\underline{e}$ & C
                                            & ($k\underline{l}$)$\underline{e}$ & C \\
                                & ($\underline{l}e$)$\underline{k}$ & C
                                      & ($l\underline{e}$)$\underline{k}$ & C
                                            &  & \\
                                & ($e\underline{k}$)$\underline{l}$ & $\sim\sim\sim$
                                      & &
                                            & ($\underline{e}k$)$\underline{l}$ & $\sim\sim\sim$ \\
      \midrule
                                & \multicolumn{2}{c}{C} & \multicolumn{2}{c}{C} & \multicolumn{2}{c}{C} \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{table}

**** For the time being, I've added two of these tables.
We can add the remaining seven tables easily, either here or in an appendix. --MKH ****

For many estimates of elasticity of substitution, the resampling technique indicates high uncertainty. 
Where results are not uncertain, 
we find complementarity among factors of production 
for most economies with the exception of China 
where an indication of substitutability between labor and energy is observed.

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \sigma_1 = 1$), 
is a troubling characteristic of the estimated CES models. 
For some countries (IR, ZM), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Note that this result is unlikely to be caused by poor-quality data alone. 
Presumably the best data in the world (US, UK, JP) exhibits large uncertainty. 
(See US and UK rows in Figures~\ref{fig:CD_Ternary_Scatter_Graph}  
and~\ref{fig:CES_sigma1_delta1_graph}--\ref{fig:CES_ternary_graph}.)

Some of the uncertainty is explained by the structure of the CES model. 
If $\delta$ or $\delta_1$ achieves an extreme value (0 or 1), 
the corresponding elasticity of substitution ($\sigma$ and $\sigma_1$) becomes meaningless. 
(See Equations~\ref{eq:CES}--\ref{eq:CESekl}.) 
Thus, large uncertainty in $\sigma$ and $\sigma_1$ is expected for any resampled model 
in which parameter estimates are positioned on the left or right boundaries 
of Figures~\ref{fig:CES_sigma1_delta1_graph} or~\ref{fig:CES_sigma_delta_graph}.

Some of the uncertainty in parameter estimates may be explained by structural change in the economies. 
For example, 
**** Martin: can you select one of the highly uncertain economies and discuss it? 
Perhaps Iran or Zambia? ****. 
To account for the large uncertainty, 
it may be useful to perform piecewise analyses of smaller time periods. 
For the present study, doing piecewise analysis of, say, 
Iran **** Use the economy that Martin discusses above **** 
would mean that the CES model is estimated on a 10-year period. 
Considering that there are six fitted parameters in the CES model 
($\gamma$, $\lambda$, $\delta$, $\delta_1$, $\rho$, and $\rho_1$), 
ten years of historical data leaves precious few (only 4) degrees of freedom for the fitting process.

Based on the results and discussions above 
regarding the relative importance of and 
the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
should be undertaken with utmost caution. 
Indeed, empirical testing on the relative importance 
of production factors and elasticities of substitution is highly recommended. 
Sensitivity analysis is essential. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies to enhance the reliability of results.

%++++++++++++++++++++++++++++++
\subsection{Implications of resampling results } 
\label{sec:implications}
%++++++++++++++++++++++++++++++

In addition to the usual parameter estimation process 
by linear coefficient determination (Cobb-Douglas and LinEx) 
and non-linear optimization algorithms (CES), 
the resampling/bootstrapping analysis performed herein 
provides insight into the question of how accurately parameters are known. 
We chose the resampling approach instead of the more-common technique 
of estimating confidence intervals from standard errors for two reasons. 
(1)~Model parameters are highly constrained, and confidence intervals often violate the constraints. 
For example a result of $\alpha = 0.25 \pm 0.3$ is nonsensical, 
because $ \alpha \in [0, 1]$. 
(2)~True confidence intervals for constrained parameters are often asymmetric
(e.g., $\alpha = 0.2 + 0.1, -0.05$), 
but the standard error approach yields symmetric confidence intervals. 
The resampling approach that we employed both respects constraints 
and allows for asymmetric confidence intervals.%
  \footnote{
    We have chosen to present parameter uncertainty results visually, 
    both because the resampling technique lends itself to visual presentation and 
    because we believe visual comparison is more effective than tablefuls of numbers.
  }
A significant result to emerge from this work is that large parameter uncertainty 
does not necessarily lead to poor model fit to economic output. 
To choose one example of many, 
the value of $\gamma$ for the Cobb-Douglas model for Tanzania (0.59) 
shows a resampling range of 0.2–0.9. 
(See Figure~\ref{fig:CD_Ternary_Scatter_Graph}.) 
Yet, Cobb-Douglas models for Tanzania, with and without energy, both 
(a)~fit historical data very well 
(white line compared to black line in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}) 
and 
(b)~exhibit a very narrow range of resampled fits 
(thin gray band in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}). 
This means that a good fit to historical data does not indicate 
that model parameters have small uncertainty. 
In fact, similar to Tanzania, 
many economies exhibit both large uncertainty in model parameters and
very good fit to historical data. 
(See Figures~\ref{fig:CD_lambda_theta_graph} ~\ref{fig:CD_Ternary_Scatter_Graph} 
compared to Figure~\ref{fig:CD_GDP_Spaghetti_Graph}
for the Cobb-Douglas models and 
Figure~\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_sigma_delta_graph}
compared to Figure~\ref{fig:CES_Spaghetti_Graph} for CES models.) 
This result is consistent across almost all economies and models. 
Thus, we conclude that great care must be taken when interpreting estimated parameters: 
it is very easy to claim more certainty than is warranted. 
Where large uncertainty is observed, 
additional work is needed to contextualize the results, 
for example, 
sector-based analysis or ****. **** Martin, add more here? ****

We further caution against using $R^2$ only as the criteria for choosing a model, 
because, in our experience, 
both Cobb-Douglas and CES models yield high $R^2$ values for all countries. 
(See Figures~\ref{fig:CD_Spaghetti_Graph} and~\ref{fig:CES_Spaghetti_Graph}.) 
LinEx models fare poorly in some cases and might be excluded on that basis. 
(See SA and TZ in Figure~\ref{fig:LINEX_Spaghetti_Graph}.) 
It is more important to know the uncertainty of estimated parameters than to assess $R^2$. 
With rigorous uncertainty testing as we have done here, 
the quality of parameter estimates is exposed: 
at a glance, the tightness of resampled results indicates 
whether or not confidence in the model and data is warranted. 
We conclude that it is insufficient to argue 
in favor of a particular production function 
on the grounds of its ability to fit historical data.

The implications of parameter uncertainties should be obvious, 
but we will state them here. 
Basing policy upon parameters with high uncertainty can lead to bad, or completely wrong, policy. 
**** Martin: can you find an example in our results that drives this point home? --MKH  
May be barking up wrong production factor, may assume subs when complementarity: ****

We conclude that the best policy will be made 
with a macroeconomic modeling approach that is highly interpretative, 
uses several production functions, and 
estimates asymmetric uncertainties of fitted parameters.
